Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-01 paper/main.tex
Updates

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused on developing
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}.

Despite the availability of many automated algorithms for constructing DAGs,
their adoption in applied research remains limited. Researchers often prefer to
manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We believe this preference persists due to several challenges
with these algorithms when applied in practical settings:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood \todo{Is there any citation for this}.
		Additionally, their output can vary significantly depending on
		the chosen algorithm and their hyperparameters. Combined with a
		lack of robust methods to test these algorithms' performance on
		a given dataset, it can be difficult to trust their outputs. As
		a result, even when using these algorithms are used,
		researchers often rely on post hoc manual modifications to the
		model based on model testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, the
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. As
		a result, researchers have to manually orient these edges based
		on their expert knowledge before using them for analysis.
\end{enumerate}

@@

There has been work on trying to come up with a single test to combine all
these test to give a single answer \citep{Eulig2023, Shipley2000}. However,
what is missing is the guide on how to how to fix these tests. Additionally
there has been a lot of work to integrate expert knowledge into causal
discovery algorithms which has shown significant improvement. To tackle these
issues with CI based testing, we take inspiration from \emph{modification
indices} in specification search. The modification indices method tries to
identify and rank changes in the model such that modifications that improve the
model fit the most are ranked higher.This ranking allows researchers to focus
first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications. Analogous
to this, in this paper we develop a measure for partial association for mixed
data that allows us to rank the CI violations. Additionally, this measure of
conditional association also allows us tackle other issues with CI based
testing outlined in the previous paragraphs.

>>>>

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. While DAG-based methods focus on automated discovery,
SEM-based methods emphasize expert-driven model specification. This includes
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}.

Despite significant progress in automated causal discovery, their adoption in
applied research has been limited. Researchers often prefer to manually
construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We attribute this preference to several challenges with existing
causal discovery algorithms in practical settings:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are asymptotically
		consistent, their behavior on finite samples is not well
		understood. Additionally, their output can vary significantly
		depending on the choice of algorithm and hyperparameters,
		making it difficult to assess reliability. The absence of
		robust performance evaluation methods further reduce the
		confidence in their outputs. As a result, even when using these
		algorithms are used, researchers often rely on post hoc manual
		modifications to the model based on model testing and their
		domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, the
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. As
		a result, researchers have to manually orient these edges based
		on their expert knowledge before using them for analysis.
\end{enumerate}

@@

% There has been work on trying to come up with a single test to combine all
% these test to give a single answer \citep{Eulig2023, Shipley2000}. However,
% what is missing is the guide on how to how to fix these tests. Additionally
% there has been a lot of work to integrate expert knowledge into causal
% discovery algorithms which has shown significant improvement.

To tackle these issues with CI based testing, we take inspiration from
\emph{modification indices} in specification search. The modification indices
method tries to identify and rank changes in the model such that modifications
that improve the model fit the most are ranked higher.This ranking allows
researchers to focus first on the suggestions that can improve the model the
most and while integrating their expertise to choose the most optimal
modifications. Analogous to this, we develop a method for ranking CI test
violations. This ranking method utilizes both p-values and a measure of
association that we develop in this paper. Using this ranking we present a
procedure for iteratively modifying the DAG while integrating expert knowledge
into it.
